Choosing the Choice: Reflections on Modelling Decisions and Behaviour in Demographic Agent-Based Models

### 2017-04-28 article.tex
jb changes

That the context in which a decision is made plays a role in the choice is inarguable: context can strongly influence the decision process itself \citep{Ben-Akiva2012}. The challenge arises in determining what the context is, and having done so, deciding what of it to consider, and how to operationalise it. The former is the easier task, since the context is the universe up to the moment of decision. The latter are more problematic, and the `state of the art' is for now, just that - an art.

@@

This iterative approach places considerable demands on the modeller in terms of developing the model, but also in exercising their judgement as to what should be included, and identifying where a component should be iterated out. This is however an expensive exercise, both because augmenting a model is time consuming, and because evaluating complex models poses considerable challenges. The challenge of evaluation can be somewhat ameliorated by the use of techniques like sensitivity analysis (see \citet{Thiele2014} for a review of several techniques, and \citet{Oakley2004} for an alternative approach), which can support the modeller in identifying how the individual parameters of a model contribute to overall outputs. A key benefit to this is that parameters which do not significantly impact the results of a model can be removed. This too has limitations, because as we discuss later in relation to heterogeneity in agents (section \ref{heterogeneity-in-decision-strategies}), context can take the form of both process, and parameter, and sensitivity analysis is directly informative only about the latter.
As a result, the onus remains on the modeller to use their best judgement. The application of such judgement can be documented, however: \cite{Schmolke2010} suggest that developmental cycles are recorded in a standardised format, the TRACE framework, so the process of rejection of alternatives through which the final model came about can be reconstructed.

@@

Neither approach is a complete account. While mimicry seems to be fundamental to many aspects of human behaviour \citep{Chartrand1999}, we are not simply impersonators and instead interpret the actions of others. Absent telepathy however, we can only draw inferences based on observation of their actions\footnote{Which may well include communicating about their internal state.}. A potential approach to this might be one similar to that of \citet{Jern2011a} who propose an inverted generative model\footnote{A generative model produces decisions, given preferences over outcomes. The inversion provides estimates of preferences, given observed decisions.} to explain how people might translate observed behaviour into preferences, although this presages a utility type mechanism for decision making. This also has implications for more high level models of the interaction between context and decision making process, for example that of \citet{Ben-Akiva2012}, where there are explicit models of the interaction between context and choice. An inverted model which does not sufficiently capture the context in which an agent's decision making took place, i.e. the observing agent is underinformed about inputs to the other's decision process, or constraints on their actions, may well lead to faulty inferences about what drove their behaviour. Although this is realistic, in that our own inferences about the motivations of others are not always accurate, it may not be desirable in all models.

@@

Time is explicitly represented in almost all agent-based models, either in discrete or (pseudo) continuous form, in order to allow the dynamic unfolding of the modelled processes by `stepping' from one time-step to the next (or between time-ordered events in continuous-time discrete-event simulation). However, time often  does not enter into the decision making process of the simulated individuals. This results in the exclusion of a number of factors that may potentially be significant in eventual model outcomes. Firstly, individuals do not always make decisions instantaneously; as \cite{Willekens2017} describe elsewhere in this special issue, the Theory of Planned Behaviour \citep{Ajzen1991} can be used to describe the gradual process of intention forming, influence-absorption, information gathering and planning that terminate in a migration event. This drawn-out decision making process allows the possibility that agents may change their minds. Secondly, past experiences affect decisions in the present. This brings up questions about how the passage of time affects memory and attitudes. Finally, agents differ in their preferences between rewards realised immediately and equivalent rewards realised at some point in the future.

@@

This assumed sensitivity of demographic decision-making to the horizons over which the resultant costs and benefits are realised suggests that we should at least acknowledge what assumptions we make about how people value present and future rewards. The flexibility of agent-based models suggests that are is also scope to examine the effect of decision models that include time preference. Considering migration models, agents may learn about the potential rewards form migration from their social environment, and such information will influence a prediction about their future conditional on their migration decision. The influence of this prediction on the final decision may then depend on the time horizon over which rewards are realised.  A brief summary of some current thinking on time-dependence in decision making follows; including such models in demographic models will certainly be challenging, and in many cases impossible, but being aware of the potential problems with making simplifying assumptions about treatment of time in decision models is valuable, and at least helps a modeller anticipate when such assumptions may fall down and why.

@@

At the level of model, there are different layers of uncertainty: in the parameters, model specification, computer code, observations, not to mention the inherent residual variability \citep{Kennedy2001}. There are different ways to treat these manifestations of uncertainty in agent-based models. The use of statistical emulators or meta-models \citep{Kleijnen2000} -- statistical models approximating the key uncertain relationships between the parameters of the underlying complex computational models -- has been suggested by \citet{Kennedy2001} and \citet{Oakley2004}, and more recently, specifically in the context of agent-based modelling, by \citet{Heard2015} and \citet{Hilton2016}. Alternatives include the use of approximate Bayesian computation, as advocated by \citep{Grazzini2017}, or hybrid approaches, whereby emulators are enhanced by direct samples from the output of the underlying agent-based models \citep{Kaminski2015}.

@@

In the context of the study of expert decision making, psychologist James Shanteau has criticised psychology for a focus on the study of the ``Generalised Normal Adult Human Mind'', while ignoring differences between people and contexts \citep{Shanteau2015}. Agent-based models provide an opportunity to study two potential ways in which individuals may differ in methods of decision-making. The first - somewhat easier to operationalise and analyse - can be termed \emph{parametric} differences, in which individuals are thought of as having the same underlying model of decision, but are different only in the way that these are parametrised. Many agent-based models include such differentiation. As an example, in the \citet{Epstein2002} model of civil disobedience, individuals differ in the extent of their aversion to risk, but for a given level of risk aversion, agents will behave in the same way (holding all else constant). Notably, this heterogeneity in risk aversion is crucial to the behaviour of the model. Without it, the `revolts' against authority which were a central feature of the model would never be triggered.

Secondly, individuals may differ in the \emph{methods} they use to come to decisions in the same situations. Sociological models regarding frames and scripts provides some justification for thinking that this may be the case \citep{Kronenberg2014}\footnote{Thanks to Sebastian Pink for drawing our attention to this strand of the literature.}. These suggest that individuals first attempt to select a frame for the situation they encounter, where are frames are mental model which answer the question: `What kind of situation is this' . Conditional on the selected frame individuals, will then look to choose a relevant script - in answer to the hypothetical question: ``What am I expected to do in this situation?'' (ibid; 99). Thus, individuals may differ widely in their behaviour should they frame the situation differently, or if their cultural experiences suggest a different script.  This second type of heterogeneity is potentially more difficult to operationalise, yet is more fundamental to the way the decision processes are described.

@@

@@

In beginning to answer the question of which model of choice behaviour is the right one, we would do well to borrow an insight from computational neuroscience, and recognise that a productive understanding of decision making in a demographic context requires understanding at multiple levels \citep{Marr1976,Marr1982}. High level models like the Theory of Planned Behaviour \citep{Ajzen1991} can assist in understanding what decisions are being made, and why, and on the sequencing of the underpinning processes. Models of process, and the integration of context like that of \citet{Ben-Akiva2012}, can inform about how decision making takes place, and what the processes involved are. Finally, these must be coupled with an appropriate implementation.

@@

So, in the light of the above arguments, how to choose the appropriate representation of the human decisions and behaviour to be used in the agent-based modelling of demographic processes? First, we need to remember that agent-based models are a very convenient tool, with which we can integrate behavioural theory with social theory and data. This implies that – as suggested in the Manifesto of computational social science \citep{Conte} – such models would ideally need to collect additional information on human decision making through bespoke cognitive experiments, instead of merely relying on hypotheses or assumptions \citep{Courgeau}. In this way, the choice of a particular choice model will become an empirical issue, but specific to a given context and the decision problem of interest. Following this approach would allow for including empirical micro-foundations in computational demographic models in an open and transparent way – and provide a natural, bottom-up and empirical way if evaluating the suitability of particular choice theories for the problem at hand.

@@

In particular, by calibrating the model, we can learn the choice of an appropriate representation of the decision processes of different agents. If the results of calibration do not point out to choosing a specific set of decision rules and models for various agents, this aspect of modelling can be enhanced by conducting cognitive experiments under controlled conditions. The sensitivity analysis would help identify, which aspects of the decision making process are the most important from the point of view of generating the observed outcomes -- this could be the treatment of time, uncertainty, or other parameters. In the fourth step of the process, the decision problems faced by the agents can be reproduced under laboratory conditions in a series of tasks to be solved by human volunteers participating in such experiments. The additional insights from this exercise would enable fine-tuning and re-parameterising of the agent-based model in the light of experimental findings, as proposed in the Manifesto of computational social science \citep{Conte}. The second, third and fourth steps above can then be iterated until no more improvements can be reasonably made -- what will be left is the residual uncertainty that is an irreducible feature of all models, and especially such complex ones.

>>>>

That the context in which a decision is made plays a role in the choice is inarguable: context can strongly influence the decision process itself \citep{Ben-Akiva2012}. The challenge arises in determining what the context is, and having done so, deciding what of it to consider, and how to operationalise it. The former is the easier task, since the context is the universe up to the moment of decision. The latter are more problematic, and the `state of the art' is for now, just that -- an art.

@@

This iterative approach places considerable demands on the modeller in terms of developing the model, but also in exercising their judgement as to what should be included, and identifying where a component should be iterated out. This is however an expensive exercise, both because augmenting a model is time consuming, and because evaluating complex models poses considerable challenges. The challenge of evaluation can be somewhat ameliorated by the use of techniques like sensitivity analysis (see \citet{Thiele2014} for a review of several techniques, and \citet{Oakley2004} for an alternative approach), which can support the modeller in identifying how the individual parameters of a model contribute to overall outputs. In this context, sensitivity analysis refers to a broad range of techniques that enable understanding, how the various model inputs affect the output (result) variables, and how important the model parameters are relative to one another \citep{Oakley2004, Thiele2014}. One important benefit to sensitivity analysis is that parameters which do not significantly impact the results of a model can be removed. This too has limitations, because as we discuss later in relation to heterogeneity in agents (section \ref{heterogeneity-in-decision-strategies}), context can take the form of both process, and parameters, and sensitivity analysis is directly informative only about the latter.As a result, the onus remains on the modeller to use their best judgement. The application of such judgement can be documented, however: \cite{Schmolke2010} suggest that developmental cycles are recorded in a standardised format, the TRACE framework, so the process of rejection of alternatives through which the final model came about can be reconstructed.

@@

Neither approach is a complete account. While mimicry seems to be fundamental to many aspects of human behaviour \citep{Chartrand1999}, we are not simply impersonators and instead interpret the actions of others. Absent telepathy however, we can only draw inferences based on observation of their actions, which may well include communicating about their internal state. A potential approach to this might be one similar to that of \citet{Jern2011a} who propose an inverted generative model to explain how people might translate observed behaviour into preferences, although this presages a utility type mechanism for decision making. Here, a generative model produces decisions, given preferences over outcomes, and the inversion provides estimates of preferences, given observed decisions. This also has implications for more high level models of the interaction between context and decision making process, for example that of \citet{Ben-Akiva2012}, where there are explicit models of the interaction between context and choice. An inverted model which does not sufficiently capture the context in which an agent's decision making took place, i.e. the observing agent is underinformed about inputs to the other's decision process, or constraints on their actions, may well lead to faulty inferences about what drove their behaviour. Although this is realistic, in that our own inferences about the motivations of others are not always accurate, it may not be desirable in all models.

@@

Time is explicitly represented in almost all agent-based models, either in discrete or (pseudo) continuous form, in order to allow the dynamic unfolding of the modelled processes by `stepping' from one time-step to the next (or between time-ordered events in continuous-time discrete-event simulation). However, time often  does not enter into the decision making process of the simulated individuals. This results in the exclusion of a number of factors that may potentially be significant in eventual model outcomes. Firstly, individuals rarely make decisions instantaneously, especially important ones with potentially far-ranging and long-lasting consequences, such as most decisions related to demographic behaviour -- entering into or ending partnerships, having children, migrating, and so on. As \cite{Willekens2017} describe elsewhere in this special issue, the Theory of Planned Behaviour \citep{Ajzen1991} can be used to describe the gradual process of intention forming, influence-absorption, information gathering and planning that terminate in a migration event. This drawn-out decision making process allows the possibility that agents may change their minds. Secondly, past experiences affect decisions in the present. This brings up questions about how the passage of time affects memory and attitudes. Finally, agents differ in their preferences between rewards realised immediately and equivalent rewards realised at some point in the future.

@@

This assumed sensitivity of demographic decision-making to the horizons over which the resultant costs and benefits are realised suggests that we should at least acknowledge what assumptions we make about how people value present and future rewards. The flexibility of agent-based models suggests that are is also scope to examine the effect of decision models that include time preference. Considering migration models, agents may learn about the potential rewards form migration from their social environment, and such information will influence a prediction about their future conditional on their migration decision. The influence of this prediction on the final decision may then depend on the time horizon over which rewards are realised.  A brief summary of some current thinking on time-dependence in decision making follows; including such tenets in demographic modelling will certainly be challenging, and in many cases impossible, but being aware of the potential problems with making simplifying assumptions about treatment of time in decision models is valuable, and at least helps a modeller anticipate when such assumptions may fall down and why.

@@

At the level of model, there are different layers of uncertainty: in the parameters, model specification, computer code, observations, not to mention the inherent residual variability \citep{Kennedy2001}. There are different ways to treat these manifestations of uncertainty in agent-based models. The use of statistical emulators or meta-models \citep{Kleijnen2000} -- statistical models approximating the key uncertain relationships between the parameters of the underlying complex computational models -- has been suggested by \citet{Kennedy2001} and \citet{Oakley2004}, and more recently, specifically in the context of agent-based modelling, by \citet{Heard2015} and \citet{Hilton2016}. Alternatives include the use of approximate Bayesian computation, as advocated by \citet{Grazzini2017}, or hybrid approaches, whereby emulators are enhanced by direct samples from the output of the underlying agent-based models \citep{Kaminski2015}.

@@

In the context of the study of expert decision making, psychologist James Shanteau has criticised psychology for a focus on the study of the ``Generalised Normal Adult Human Mind'', while ignoring differences between people and contexts \citep{Shanteau2015}. Agent-based models provide an opportunity to study two potential ways in which individuals differ in methods of decision-making. The first -- somewhat easier to operationalise and analyse -- can be termed \emph{parametric} differences, in which individuals are thought of as having the same underlying model of decision, but are different only in the way that these are parametrised. Many agent-based models include such differentiation. As an example, in the \citet{Epstein2002} model of civil disobedience, individuals differ in the extent of their aversion to risk, but for a given level of risk aversion, agents will behave in the same way (holding all else constant). Notably, this heterogeneity in risk aversion is crucial to the behaviour of the model. Without it, the `revolts' against authority which were a central feature of the model would never be triggered.

Secondly, individuals may differ in the \emph{methods} they use to come to decisions in the same situations. Sociological models regarding frames and scripts provides some justification for thinking that this may be the case \citep{Kronenberg2014}. These suggest that individuals first attempt to select a frame for the situation they encounter, where are frames are mental model which answer the question: `What kind of situation is this' . Conditional on the selected frame individuals, will then look to choose a relevant script - in answer to the hypothetical question: ``What am I expected to do in this situation?'' (ibid; 99). Thus, individuals may differ widely in their behaviour should they frame the situation differently, or if their cultural experiences suggest a different script.  This second type of heterogeneity is potentially more difficult to operationalise, yet is more fundamental to the way the decision processes are described.

@@

A systematic review of various aspects of decision making in agent-based models of land use has been recently published by \citet{Groeneveld}. The review looks at some of the aspects discussed at this paper, such as uncertainty or heterogeneity, at very a general level, and provides useful insights into the current modelling practice in a specific area of application. In particular, in the review by \citet{Groeneveld}, only 6 out of 118 models studied were found to include some uncertainty analysis of the decision model, and only 17 out of 128 have incorporated uncertainty in the decision processes of agents. There is certainly room for further improvement in that respect, and this conclusion holds for demographic models as much as they do for the studies of land use.

@@

In beginning to answer the question of which model of choice behaviour is the right one, we would do well to borrow an insight from computational neuroscience, and recognise that a productive understanding of decision making in a demographic context requires understanding at multiple levels of theoretical abstraction or generality \citep{Marr1976,Marr1982}. High-level models like the Theory of Planned Behaviour \citep{Ajzen1991} can assist in understanding what decisions are being made, and why, and on the sequencing of the underpinning processes. Models of process, and the integration of context like that of \citet{Ben-Akiva2012}, can inform about how decision making takes place, and what the processes involved are. Finally, these must be coupled with an appropriate implementation.

@@

So, in the light of the above arguments, how to choose the appropriate representation of the human decisions and behaviour to be used in the agent-based modelling of demographic processes? First, we need to remember that agent-based models are a very convenient tool, with which we can integrate behavioural theory with social theory and data. This implies that -- as suggested in the Manifesto of computational social science \citep{Conte} -- such models would ideally need to collect additional information on human decision making through bespoke cognitive experiments, instead of merely relying on hypotheses or assumptions \citep{Courgeau}.

In this way, the choice of a particular choice model and the underlying choice theory will become largely an empirical issue, being driven to a great extent by the empirical observations and experimental data on decision making in a particular context, and specific to the decision problem of interest. Following this approach would allow for including empirical micro-foundations in computational demographic models in an open and transparent way -- and provide a natural, bottom-up and empirical way if evaluating the suitability of particular choice theories for the problem at hand. Such a methodology would also take forward and expand on some of the suggestions mentioned by \citet{Groeneveld}, including a call for a fuller use of knowledge on human decision making and greater cross-disciplinary collaboration in modelling.

@@

In particular, by calibrating the model, we can learn the choice of an appropriate representation of the decision processes of different agents. If the results of calibration do not point out to choosing a specific set of decision rules and models for various agents, this aspect of modelling can be enhanced by conducting cognitive experiments under controlled conditions. The sensitivity analysis would help identify, which aspects of the decision making process are the most important from the point of view of generating the observed outcomes -- this could be the treatment of time, uncertainty, or other parameters.

In the fourth step of the process, the decision problems faced by the agents can be reproduced under laboratory conditions in a series of tasks to be solved by human volunteers participating in such experiments. The experiments would be designed in such a way, so as to structurally correspond to the decision problem at hand. For example, in the context of the neoclassical theory of migration, the participants could decide whether or not to play a lottery game, whereby they would either lose money instantaneously (corresponding to an unsuccessful migration outcome), or win a promise of a monetary payment at some time in the future (corresponding to a successful one). The decisions of participants on whether or not to play the game could then be operationalised in the simulated decision making of the modelled agents by using the particular decision theory that is found most closely corresponding to the patterns observed in the experiment.

The additional insights from the experimental data collection exercise would enable fine-tuning and re-parameterising of the agent-based model in the light of the additional findings, as proposed in the Manifesto of computational social science \citep{Conte}. The second, third and fourth steps above can then be iterated until no more improvements can be reasonably made -- what will be left is the residual uncertainty that is an irreducible feature of all models, and especially such complex ones.
